---
title: "Data Visualization in R"
format: html
editor: visual
---

## Static Visualization: ggplot2
```{r}
# install.packages("library_name")
library(ggplot2)
data(Theoph)
```

### ggplot()
ggplot() is the foundation of generating a plot in ggplot environment.

However, the code below won't show any plot because we haven't defined any type of plots we want. 
```{r}
ggplot(Theoph, aes(x = Time, y = conc))
```

### geom_*()

We can use geom_*() to define geometric objects that can be placed on a graph. 

Example: scatterplot of drug concentration over time.
```{r}
ggplot(Theoph, aes(x = Time, y = conc)) +
  geom_point()
```

#### Add color, size, and alpha
color - change the color of the plot
alpha - transparency ranges from 0 (completely transparent) to 1 (completely opaque)
size - size of the dot
```{r}
ggplot(Theoph, aes(x = Time, y = conc)) +
  geom_point(color = "steelblue", ## change the color of the dots
             alpha = 0.7, ## transparency
             size = 2 ) ## size of the dots
```

### geom_smooth()

This function adds a line of best fit of the graph.
```{r}
ggplot(Theoph, aes(x = Time, y = conc)) +
  geom_point(color = "steelblue", alpha = 0.6, size = 2) +
  geom_smooth()
```

#### Adjust line types, thickness, color, and confidence interval

method - different line types. Common options: loess, lm, etc.
color - change color for the line
linewidth - thickness of the line
se - confidence band around smooth curve

```{r}
ggplot(Theoph, aes(x = Time, y = conc)) +
  geom_point(color = "steelblue", 
             alpha = 0.6, 
             size = 2) +
  geom_smooth(method = "lm", ## Line types
              color = "red", ## Color
              linewidth = 1.5, ## thickness
              se = TRUE) ## confidence interval
```
### labs()

The label function provides customized labels for the axes and legends. 

title - title of the plot
subtitle - subtitle of the plot (optional)
x - x-axis label
y - y-axis label

```{r}
ggplot(Theoph, aes(x = Time, y = conc)) +
  geom_point(color = "steelblue", alpha = 0.6, size = 2) +
  geom_smooth(se = FALSE) +
  labs(title = "Drug concentration over time",
       x = "Time since dose (hours)",
       y = "Concentration") 
```

#### finish the graph with a theme

theme_*() has a lot of different choices:
- theme_light
- theme_dark
- theme_classic
...

```{r}
ggplot(Theoph, aes(x = Time, y = conc)) +
  geom_point(color = "steelblue", 
             alpha = 0.6, 
             size = 2) +
  geom_smooth(se = FALSE) +
  labs(title = "Drug concentration over time",
       x = "Time since dose (hours)",
       y = "Concentration") +
  theme_minimal() ## A minimalistic theme with no background annotations.
```

### Grouping
```{r}
## divide weight into two groups
Theoph$WtGroup <- ifelse(Theoph$Wt >= median(Theoph$Wt), "Higher weight", "Lower weight")

## use weight as color separation in the ggplot() function
ggplot(Theoph, aes(x = Time, y = conc, color = WtGroup)) +
  geom_point(alpha = 0.7)
```

We can tell that there are now two lines on the graph now.
```{r}
## use weight as color separation in the ggplot() function
ggplot(Theoph, aes(x = Time, y = conc, color = WtGroup)) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE)
```

#### Special case: Spaghetti plots
```{r}
ggplot(Theoph, aes(x = Time, y = conc)) +
  geom_line()
```

```{r}
ggplot(Theoph, aes(x = Time, y = conc, group = Subject)) +
  geom_line()
```

### Scales
Scales control how values appear on the axes and in the plot. 

Going back to the spaghetti plot, we can adjust the axes and make the plot looks more fit
```{r}
ggplot(Theoph, aes(Time, conc, group = Subject)) +
  geom_line(alpha = 0.3, 
            color = "gray50") +
  geom_smooth(aes(group = 1), 
              se = FALSE, 
              color = "red")
```

Check the range of axes
```{r}
range(Theoph$Time)
range(Theoph$conc)
```

Adjust the x & y range
```{r}
ggplot(Theoph, aes(Time, conc, group = Subject)) +
  geom_line(alpha = 0.3, 
            color = "gray50") +
  geom_smooth(aes(group = 1), 
              se = FALSE, 
              color = "red") +
  scale_x_continuous(breaks = seq(0, 25, 5),
                     labels = paste0(seq(0, 25, 5), "hr")) +
  scale_y_continuous(breaks = seq(0, 12, 2),
                     labels = paste0(seq(0, 12, 2), "mg/L"))
```

Add the labs
```{r}
ggplot(Theoph, aes(Time, conc, group = Subject)) +
  geom_line(alpha = 0.3, 
            color = "gray50") +
  geom_smooth(aes(group = 1), 
              se = FALSE, 
              color = "red") +
  scale_x_continuous(breaks = seq(0, 25, 5)) +
  scale_y_continuous(breaks = seq(0, 12, 2)) +
  labs(title = "Drug concentration over time",
       x = "Time (hours)",
       y = "Concentration") +
  theme_minimal()
```

### Facet
Facets split a plot into multiple panels to compare groups. 

```{r}
unique(Theoph$WtGroup)
```

What if we add weight group as the comparison groups?
```{r}
ggplot(Theoph, aes(Time, conc, group = Subject)) +
  geom_line(alpha = 0.3, 
            color = "gray50") +
  geom_smooth(aes(group = 1), 
              se = FALSE, 
              color = "red") +
  facet_wrap(~ WtGroup) +
  scale_x_continuous(breaks = seq(0, 25, 5)) +
  scale_y_continuous(breaks = seq(0, 12, 2)) +
  labs(title = "Drug concentration over time",
       x = "Time (hours)",
       y = "Concentration") +
  theme_minimal()
```
Now the plots split into two panels corresponding to the groups we defined. 

#### Other Facet function
facet_grid() 
- takes two variables
- one define rows and one defines column. 
```{r}
ggplot(Theoph, aes(Time, conc, group = Subject)) +
  geom_line(alpha = 0.3) +
  facet_grid(WtGroup ~ Subject)
```

facet_null() 
- no faceting
- default behavior
```{r}
ggplot(Theoph, aes(Time, conc, group = Subject)) +
  geom_line(alpha = 0.3) +
  facet_null()
```

### Univariate Graphs
#### Categorical 
```{r}
# install.packages("mosaicData")
data(Marriage, package = "mosaicData")
```

##### Bar Chart
Plot the distribution of race
```{r}
ggplot(Marriage, aes(x = race)) + 
  geom_bar()
```

Use ?geom_bar() to see the options we can tune for the graph. Then set the fill as "cornflowerblue" and color as "black".

```{r}
ggplot(Marriage, aes(x=race)) + 
  geom_bar(fill = "cornflowerblue", 
           color="black")
```

Add labels to the graph to match the final graph on powerpoint
```{r}
ggplot(Marriage, aes(x=race)) + 
  geom_bar(fill = "cornflowerblue", 
           color="black") +
  labs(title = "Participants by race",
       x = "Race", 
       y = "Frequency")
```

We can sort the categories by frequency
```{r}
library(dplyr)

## Get the count of each race
plotdata <- Marriage %>%
 count(race)

## Use reorder to plot the bars in ascending order
ggplot(plotdata, aes(x = reorder(race, n), y = n)) + 
  geom_bar(fill = "cornflowerblue", 
           color="black",
           stat="identity") +
  labs(title  = "Participants by race",
       x = "Race", 
       y = "Frequency")
```

```{r}
## Use reorder to plot the bars in descending order
ggplot(plotdata, aes(x = reorder(race, -n), y = n)) + 
  geom_bar(fill = "cornflowerblue", 
           color="black",
           stat="identity") +
  labs(title  = "Participants by race",
       x = "Race", 
       y = "Frequency")
```

We can also add count on top of each bar to make the graph clearer

geom_text adds the labels, and vjust controls vertical justification. 
```{r}
# plot the bars with numeric labels
ggplot(plotdata, aes(x = race, y = n)) + 
  geom_bar(fill = "cornflowerblue", 
           color="black",
           stat="identity") +
  geom_text(aes(label = n), vjust=-0.5) +
  labs(title  = "Participants by race",
       x = "Race", 
       y = "Frequency")
```

Sometimes the plot could have overlapping labels
```{r}
ggplot(Marriage, aes(x=officialTitle)) + 
  geom_bar() +
  labs(title = "Marriages by officiate",
       x = "Officiate",
       y = "Frequency")
```

We can use the coord_flip() function to flip the x and y axes.
```{r}
ggplot(Marriage, aes(x = officialTitle)) + 
  geom_bar() +
  labs(title = "Marriages by officiate",
       x = "",
       y = "Frequency") +
  coord_flip()
```

Or we could also rotate the axis labels
```{r}
ggplot(Marriage, aes(x=officialTitle)) + 
  geom_bar() +
  labs(x = "",
       y = "Frequency",
       title = "Marriages by officiate") +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1))
```

#### Quantitative 
##### Histogram
Plot the distribution of age using histogram
```{r}
ggplot(Marriage, aes(x = age)) +
  geom_histogram()
```

2. Use ?geom_histogram() to see the options we can tune for the graph. 

- Set the fill as "cornflowerblue" and color as ”white”
- Adjust the bins to be 20
```{r}
ggplot(Marriage, aes(x = age)) +
  geom_histogram(fill = "cornflowerblue", 
                 color = "white",
                 bins = 20)
```

3. Add labels to the graph to match the final graph shown on the powerpoint
```{r}
ggplot(Marriage, aes(x = age)) +
  geom_histogram(fill = "cornflowerblue", 
                 color = "white",
                 bins = 20) + 
  labs(title="Participants by age",
       x = "Age")
```

Alternative way of "bins": binwidth
```{r}
ggplot(Marriage, aes(x = age)) +
  geom_histogram(fill = "cornflowerblue", 
                 color = "white", 
                 binwidth = 5) + 
  labs(title="Participants by age", 
       x = "Age")
```

##### Density plot
We can use a density plot instead of represent quantitative data. 
```{r}
ggplot(Marriage, aes(x = age)) +
  geom_density() + 
  labs(title = "Participants by age")
```

If we want to highlight the area under the curve, we can specify it in the fill option
```{r}
ggplot(Marriage, aes(x = age)) +
  geom_density(fill = "indianred3") + 
  labs(title = "Participants by age")
```

### Bivariate Graphs
#### Categorical vs. Categorical
```{r}
Theoph$WtGroup <- ifelse(Theoph$Wt >= median(Theoph$Wt),
                         "High weight", "Low weight")

Theoph$TimeGroup <- ifelse(Theoph$Time <= 10,
                           "Early", "Late")
```

##### Stacked Bar
Pros:
- It shows total and composition at the same time
- efficient on space
- good for showing part-to-whole relationship

Cons:
- Hard to compare segements across categories
- Only the bottom segment is easy to compare
- Can become cluttered with many groups
```{r}
ggplot(Theoph, aes(x = WtGroup, fill = TimeGroup)) +
  geom_bar(position = "stack")
```

##### Grouped Bar
Pros:
- Best for comparing values across groups
- Easy to read and interpret
- Clear separation between categories

Cons:
- Doesn't sohw total directly
- Takes more space (more bars if more categories)

```{r}
ggplot(Theoph, aes(x = WtGroup, fill = TimeGroup)) +
  geom_bar(position = "dodge")
```

##### Segmented bar
Pros: 
- Clearly shows proportions/percentages
- Easy to compare relative composition
- All bars are on the same scale (0-100%)

Cons:
- Loses information about absolute totals
- Hard to compare segment sizes
- Can be misleading if totals matter
```{r}
ggplot(Theoph, aes(x = WtGroup, fill = TimeGroup)) +
  geom_bar(position = "fill")
```

#### Quantitative vs. Quantitative
##### Scatterplot
It's the very first example we've done. 
```{r}
ggplot(Theoph, aes(x = Time, y = conc)) +
  geom_point(color = "steelblue", 
             alpha = 0.7, 
             size = 2 )
```

##### Line Plot
Spaghetti plot is a special line plot
```{r}
ggplot(Theoph, aes(Time, conc, group = Subject)) +
  geom_line(alpha = 0.3)
```

#### Categorical vs Quantitative
##### Box Plot
```{r}
ggplot(Theoph, aes(WtGroup, conc)) +
  geom_boxplot()
```

##### Violin Plot
```{r}
ggplot(Theoph, aes(WtGroup, conc)) +
  geom_violin()
```

## Interactive Visualization
### ggplotly()
```{r}
library(plotly)

p <- ggplot(Theoph, aes(x = Time, y = conc)) +
  geom_point(color = "steelblue", 
             alpha = 0.6, 
             size = 2) +
  geom_smooth(method = "loess", 
              color = "red",
              linewidth = 1.5,
              se = FALSE) +
  labs(title = "Drug concentration over time",
       x = "Time since dose (hours)",
       y = "Concentration")

p
ggplotly(p)
```

### plot_ly()
#### Boxplot
```{r}
plot_ly(data = Theoph, x = ~WtGroup, y = ~conc, type = "box")
```

#### Scatterplot
Plotly
```{r}
plot_ly(Theoph, x = ~Time, y = ~conc,
        type = "scatter", mode = "markers", color = ~WtGroup,
        text = ~paste("Subject:", Subject,
                      "<br>Time:", Time,
                      "<br>Concentration:", conc,
                      "<br>Weight:", Wt),
        hoverinfo = "text") |> 
  layout(title = "Scatterplot of Concentration and Time between Two weight groups",
         xaxis = list(title = "Time"),
         yaxis = list(title = "Concentration"))
```

If you would like to change color of the groups
```{r}
plot_ly(Theoph, x = ~Time, y = ~conc,
        type = "scatter", mode = "markers", color = ~WtGroup,
        colors = c("blue", "red"),
        text = ~paste("Subject:", Subject,
                      "<br>Time:", Time,
                      "<br>Concentration:", conc,
                      "<br>Weight:", Wt),
        hoverinfo = "text") |> 
  layout(title = "Scatterplot of Concentration and Time between Two weight groups",
         xaxis = list(title = "Time"),
         yaxis = list(title = "Concentration"))
```

#### Advanced Plotly Scatterplot with Animation
```{r}
theoph_anim <- Theoph %>%
  plot_ly(x = ~Time, y = ~conc, frame = ~-Subject,
          text = ~paste("Subject:", Subject,
                        "<br>Time:", Time,
                        "<br>Concentration:", conc),
    hoverinfo = "text",
    type = "scatter",
    mode = "markers") |> 
  layout(title = "Animated Drug Concentration Plot",
         xaxis = list(title = "Time"),
         yaxis = list(title = "Concentration"),
         showlegend = FALSE) ## Remove legend 

theoph_anim
```


If we want to slow down the pacing, we can customize our options
- 1000 is 1000 milliseconds per frame
- easing = linear: how the transition happens
    - linear is constant speed
- redraw = FALSE: don't redraw the entire plot every frame, just update the changes

```{r}
theoph_anim2 <- theoph_anim %>%
  animation_opts(1000, easing = "linear", redraw = FALSE)

theoph_anim2
```

#### Advanced Plotly Scatterplot with Filtering
If we only want to keep the observations below the average concentration, we can use the transforms function to help us filter out inside the plot. 
```{r}
plot_ly(data = Theoph, x = ~Time, y = ~conc, 
        type = "scatter", text = ~paste("Subject:", Subject,
                                        "<br>Time:", Time,
                                        "<br>Concentration:", conc),
        hoverinfo = "text",
        mode = "markers",
        transforms = list(list(type = "filter",
                               target = ~conc,
                               operation = "<",
                               value = mean(Theoph$conc))))
```

### Try it yourself!
```{r}
data(mtcars)
```

Create a scatterplot of weight vs mpg
```{r}
plot_ly(x = ~wt, y = ~mpg, data = mtcars,
        type = "scatter", mode = "markers")
```

Color the points by number of cylinders (cyl)
```{r}
plot_ly(x = ~wt, y = ~mpg, data = mtcars,
        color = ~cyl, type = "scatter", mode = "markers")
```

Add custom hover
```{r}
plot_ly(x = ~wt, y = ~mpg, data = mtcars,
        color = ~cyl, text = ~paste("MPG:", mpg,
                                    "<br>Weight:", wt,
                                    "<br>Cylinders:", cyl),
        hoverinfo = "text",
        type = "scatter",
        mode = "markers")
```

Add horsepower (hp) to the hover text
```{r}
plot_ly(x = ~wt, y = ~mpg, data = mtcars,
        color = ~cyl, text = ~paste("MPG:", mpg,
                                    "<br>Weight:", wt,
                                    "<br>Cylinders:", cyl,
                                    "<br>HP:", hp),
        hoverinfo = "text",
        type = "scatter",
        mode = "markers")
```


